OPC in R GIS
========================================================

These R scripts use GIS packages in R to calculate morphological complexity measure OPC from raster grid text files.

Initialise R:

```{r Initialise}
require(raster)
require(igraph)
```


GIS in R
-------------------------

Define filename and read file into matrix. Convert matrix into raster object using `rasterFromXYZ()` function. Add projection method for raster object, as this is required for hillShade plot (just keep this default).


```{r GIS_in_R, cache=TRUE}
setwd("C:/Users/arevans/Google Drive/Work desktop/0Current work/R/CatsMiceOPCDataset") #laptop
Filename <- "Felis_silvestris-1.780-uw_lrm1bemnewor150b.dat"
InputG <- read.table(Filename, header=FALSE, sep=",", na.strings="NA", dec=".", strip.white=TRUE)
RasterG <- rasterFromXYZ(InputG)
projection(RasterG) <- "+proj=eqc +lat_ts=0"
```

Create some initial plots of the raster object.

```{r GIS_Plot, cache=TRUE}
plot(x = RasterG, col = grey(0:100/100))
plot(x = RasterG, col = terrain.colors(255))
```

Calculate slope and aspect maps using `terrain` function. Use these maps to create a hillShade object, then plot it and overlay it with a topographic map. 

```{r GIS_hillShade, cache=TRUE}
RasterS <- terrain(RasterG, opt='slope')
RasterA <- terrain(RasterG, opt='aspect')
RasterH <- hillShade(RasterS, RasterA, 35, 335)
plot(RasterH, col=grey(0:100/100), legend=FALSE)
plot(RasterG, col=rev(terrain.colors(255, alpha=0.35)), add=TRUE)
```

Dental Complexity
=====================

Reclassify this aspect map using the `reclassify` function so that orientations from 0 to 22.5ยบ and 337.5 to 360ยบ are 1 (north), 22.5 to 67.5ยบ are 2(north-east), etc. The function `Define_rcl` creates the reclassification matrix for `NoCat` categories (default = 8) needed as input for the `reclassify` function. 

```{r rcl, cache=TRUE}
Define_rcl <- function (NoCat = 8) {
  rcl <- matrix(nrow = NoCat+1, ncol = 3)
  CatSize <- 360 / NoCat
  rcl[,1] <- c(0,seq(from = CatSize/2, to = 360-(CatSize/2), by = CatSize))
  rcl[,2] <- c(seq(from = CatSize/2, to = 360-(CatSize/2), by = CatSize), 360)
  rcl[,3] <- c(1:NoCat,1)
  result <- rcl
}

rcl <- Define_rcl(8)
rcl
```

Create raster `RasterAD` that shows aspect of surface in degrees, then reclassify it based on `rcl` matrix. Plot the aspect degrees and reclassified aspect maps.

```{r Aspect_Plot, cache=TRUE}
RasterAD <- terrain(RasterG, opt='aspect', unit='degrees')

plot(RasterAD, col=heat.colors(255))

RasterADC <- reclassify(RasterAD,rcl)

aspect.colors <- colorRampPalette(c("yellow","lightblue","blue","lightgreen","darkgreen","purple", "red", "brown"), space = "rgb")
plot(RasterADC, col=aspect.colors(255))
```

The aspect map of each direction (e.g. north) can be clumped to show all discrete patches of surface that face the same direction. The `freq` function shows the size (in pixels) of each discrete patch for that direction.

```{r Clump, cache=TRUE}
RasterC1 <- clump(RasterADC==1,directions=4,gaps=FALSE)
plot(RasterC1)
freq(RasterC1)
```

The `ClumpAspect` function calculates the clump map for all `NoCat` directions and creates a raster stack as output.

```{r ClumpAspect, cache=TRUE}
ClumpAspect <- function(RasterADC, NoCat) {
  result <- 0
  for (i in 1:NoCat) {
    RasterClump <- clump(RasterADC==i,directions=4,gaps=FALSE)
    
    if(i==1) { RasterClumpStack <- RasterClump }
    else { RasterClumpStack <- stack(RasterClumpStack,RasterClump) }
  }
  names(RasterClumpStack) <- c(paste(rep("clumps",NoCat),1:NoCat, sep = ""))
  result <- RasterClumpStack
}

NoCat <- 8
RasterC <- ClumpAspect(RasterADC, NoCat)
plot(RasterC)
```

The `CompileClumpData` function calculates the sizes of all clumps for the stack of clump raster maps and outputs a data frame with all of the raw values.

```{r CombileClumpData, cache=TRUE}
# Must modify to allow for NoCat != 8
CompileClumpData <- function(ClumpsOutput) {
  result <- 0
  ClumpsFreq <- freq(RasterC)
  ClumpsFreq <- freq(ClumpsOutput)
  NoCat <- length(ClumpsFreq)
  for(i in 1:NoCat) {
    ClumpsFreq[[i]] <- cbind(c(rep(i,(length(ClumpsFreq[[i]])/2))),ClumpsFreq[[i]])
  }
  ClumpResults <- rbind(ClumpsFreq[[1]], ClumpsFreq[[2]])
  for(j in 3:NoCat) {
    ClumpResults <- rbind(ClumpResults, ClumpsFreq[[j]])
  }
  colnames(ClumpResults)[1] <- "Cat"  
  result <- data.frame(ClumpResults)
}

ClumpResults <- CompileClumpData(RasterC)
```

The number of aspect clumps, or orientation patch count (OPC) of this map can be calculated. Only patches above a minimum size of `MinClumpSize` are counted.

```{r OPC_single_raster, cache=TRUE}
MinClumpSize <- 2 #Only count clumps larger than MinClumpSize pixels in size
OPCClumps <- subset(ClumpResults, ClumpResults$count > MinClumpSize)
OPCClumps <- na.omit(OPCClumps)
OPC <- nrow(OPCClumps)

OPC
```

The summary statistics for patch size are calculated using the `OPCStats` function.

```{r OPC_stats, cache=TRUE}
OPCStats <- t(as.matrix(summary(OPCClumps$count)))
OPCStats <- as.data.frame(OPCStats)
OPCStats$SD <- sd(OPCClumps$count)                          
OPCStats <- rbind(c(OPC,OPCStats))
colnames(OPCStats)[1] <- "OPC"

OPCStats
```

OPC Function
-----------------

The function `OPCCalc` takes a topographic raster map and carries out OPC calculation, outputting OPC statistics.

```{r OPCCalc, cache=TRUE}
OPCCalc <- function(RasterG, NoCat=8, MinClumpSize=2) {
  RasterAD <- terrain(RasterG, opt='aspect',unit='degrees')
  rcl <- Define_rcl(NoCat)
  RasterADC <- reclassify(RasterAD,rcl)
  RasterC <- ClumpAspect(RasterADC,NoCat)
  ClumpResults <- CompileClumpData(RasterC)

  OPCClumps <- subset(ClumpResults, ClumpResults$count > MinClumpSize)
  OPCClumps <- na.omit(OPCClumps)
  OPC <- nrow(OPCClumps)
  OPCData <- as.data.frame(OPC)
  
  OPCStats <- t(as.matrix(summary(OPCClumps$count)))
  OPCStats <- as.data.frame(OPCStats)
  OPCStats$SD <- sd(OPCClumps$count)                          
  OPCStats <- merge(OPCData,OPCStats)
  return(OPCStats)
}

OPCCalc(RasterG)
```

OPCR Function
--------------

The calculation of morphological complexity can be repeated to take into account differences in orientation of the grid around the vertical z axis, i.e. if the grid were rotated around the vertical axis, the boundaries of the directions would differ. OPCR is the mean of the OPC calculation repeated `NoRot` times. 

```{r OPCRCalc, cache=TRUE}
Define_rclR <- function (NoCat, NoRot, CurrRot) {
  rcl <- matrix(nrow = NoCat+1, ncol = 3)
  CatSize <- 360 / NoCat
  OffsetSize <- CatSize / NoRot
  Offset <- (CurrRot - 1) * OffsetSize
  
  if (CurrRot <= NoRot/2 + 1) {
    rcl[,1] <- c(0,seq(from = (CatSize/2) + Offset, to = 360-(CatSize/2) + Offset, by = CatSize))
    rcl[,2] <- c(seq(from = (CatSize/2) + Offset, to = 360-(CatSize/2) + Offset, by = CatSize), 360)
    rcl[,3] <- c(1:NoCat,1) 
  }
  else {
    rcl[,1] <- c(seq(from = Offset - (CatSize/2), to = 360, by = CatSize),0)
    rcl[,2] <- c(seq(from = (CatSize/2) + Offset, to = 360, by = CatSize), 360, Offset - CatSize/2)
    rcl[,3] <- c(1:NoCat,8)
  }
  result <- rcl
}

OPCRCalc <- function(RasterGrid, NoCat=8, MinClumpSize=2, NoRot=8, Plot=FALSE) {
  RasterAspectDeg <- terrain(RasterGrid, opt='aspect',unit='degrees')

  Rots <- c(1:NoRot)
  OPCResultsRot <- data.frame(Rots, "OPC"=NA, "Min."=NA ,"1st Qu."=NA,"Median"=NA,"Mean"=NA,"3rd Qu."=NA,"Max."=NA,"SD"=NA)
  
  CatSize <- 360 / NoCat
  
  for (i in 1:NoRot) {
    rclR <- Define_rclR(NoCat,NoRot,i)
    RasterAspectDegClass <- reclassify(RasterAspectDeg,rclR)
    if (Plot==TRUE) { plot(RasterAspectDegClass) }
    RasterClumped <- ClumpAspect(RasterAspectDegClass,NoCat)
  
    ClumpResults <- CompileClumpData(RasterClumped)
  
    OPCClumps <- subset(ClumpResults, ClumpResults$count > MinClumpSize)
    OPCClumps <- na.omit(OPCClumps)
    OPC <- nrow(OPCClumps)
    OPCData <- as.data.frame(OPC)
    
    OPCStats <- t(as.matrix(summary(OPCClumps$count)))
    OPCStats <- as.data.frame(OPCStats)
    OPCStats$SD <- sd(OPCClumps$count)                          
    OPCStats <- merge(OPCData,OPCStats)

    OPCResultsRot[i,1] <- i
    OPCResultsRot[i,2:9] <- OPCStats
    
  }
  OPCRStats <- apply(OPCResultsRot, 2, mean)
  OPCRStats <- OPCRStats[2:9]
  return(OPCRStats)
}

OPCRCalc(RasterG)
```


OPCR for All Files in Directory
----------------------------

The OPCR procedure can be carried out for all files in the current working directory. The results will be compiled into a dataframe.

```{r RunOPCRDir, cache=TRUE}
setwd("C:/Users/arevans/Google Drive/Work desktop/0Current work/R/CatsMiceOPCDataset") #laptop
RunOPCRDir <- function(NoCat=8, MinClumpSize=2, NoRot=8, MaxFiles=0, Plot=FALSE) {
  Filenames <- list.files(pattern = "\\.dat$")
  if (MaxFiles == 0) {
    NoFiles <- length(Filenames)
  }
  else {
    NoFiles <- MaxFiles
  }
  OPCRResultsFiles <- data.frame(Filenames, "OPCR"=NA, "Min."=NA ,"1st Qu."=NA,"Median"=NA,"Mean"=NA,"3rd Qu."=NA,"Max."=NA,"SD"=NA)
  result <- 0
  for (i in 1:NoFiles) {
    InputGrid <- read.table(Filenames[i], header=FALSE, sep=",", na.strings="NA", dec=".", strip.white=TRUE)
    RasterGrid <- rasterFromXYZ(InputGrid)
    projection(RasterGrid) <- "+proj=eqc +lat_ts=0"
    
    OPCRStats <- OPCRCalc(RasterGrid, Plot=Plot)
    
    OPCRResultsFiles[i,2:9] <- OPCRStats
  }
  
  result <- OPCRResultsFiles  
}

OPCRResults <- RunOPCRDir(MaxFiles = 4)
OPCRResults <- na.omit(OPCRResults)

OPCRResults
```
